---
title: "Numbers, Reciprocals, and Averages"
author: "Julio Huato"
date: "5/9/2020"
output: html_document
---

Introduction
============

This document reviews numerical notation and important arithmetic
operations regularly used in economics courses: the reciprocal or
multiplicative inverse, and two types averages: (1) the arithmetic or
simple mean and (2) the weighted mean. There will also be a tiny bit of
algebra below. Just a tiny bit.

Numerical notation
==================

Economic data appears in different but essentially equivalent numerical
forms. For certain purposes, a certain numerical notation is preferred
to others or is more convenient than others. This is a matter
ofconvenience, convention, and habit.

Here are some examples:

**Decimals** The Bureau of Economic Analysis estimated the *gross
domestic product* of the United States in 2011 at 14.99 trillion
dollars.

**Percentages** The Bureau of Labor Statistics estimated that, as of
March 2013, the U.S. *unemployment rate* was 7.6%.

**Fractions or ratios** The Bureau of Economic Analysis also estimated
that U.S. *imports* of goods and services in 2007 were approximately 1/6
of gross domestic product. Alternatively, the ratio of U.S. imports of
goods and services to GDP in 2007 was 1:6.

**Fractions or ratios** The Census Bureau reported that in 2007, the
ratio of money income of the poorest fifth of U.S. households to the
richest fifth was approximately 1:15, the highest of all developed
economies. Alternatively, the money income of the poorest fifth of U.S.
households was approximately 1/15 of that of the richest fifth.

Students of macroeconomics should visit and get well acquainted with the
following sources of U.S. economic data:

-   **Bureau of Economic Analysis**: <https://bea.gov/>

-   **Bureau of Labor Statistics**: <https://www.bls.gov/>

-   **Federal Reserve**: <https://www.federalreserve.gov/>

-   **Census Bureau**: <https://census.gov/>

-   **FRED**: <https://fred.stlouisfed.org/>

In reading and manipulating numbers, one should be able to go back and
forth between them. The following examples show the equivalences:

$$20 = 20.00 = \frac{20}{1} = \frac{200}{10} = \frac{2000}{100} = \dots = 20:1 = 2000\%.$$
$$1 = 1.0 = \frac{1}{1} = \frac{5}{5} = \frac{777}{777} = \dots = 1:1 = 100\%. $$
$$0.5 = \frac{1}{2} = \frac{2}{4} = \frac{13}{26} = \dots = 1:2 = 50\%. $$
$$0.333\ldots = \frac{1}{3} = \frac{300}{900} = \dots = 1:3 = 33.\bar{3}\%. $$
$$0.25 = \frac{1}{4} = \frac{10}{40} = \frac{25}{100} = \dots = 1:4 = 25\%.$$

Note that a number can be correctly expressed in fractional notation in
an infinite number of ways, as long as the proportion between numerator
and denominator is kept.

Reciprocals
===========

One often needs to take reciprocals (also called "multiplicative
inverses"). The **reciprocal** of a number $x$ is that number divided
into 1 or 1/$x$.

These are different forms to express the reciprocals of 50, 5, 0.5, and
0.01, respectively: 
$$\frac{1}{50} = 1:50 = 2\% $$
$$\frac{1}{5} = 1:5 = 20\% $$
$$\frac{1}{0.5} = 2 = 200\% $$
$$\frac{1}{0.01} = 100 = 10,000\%.$$

Note that when one divides any number by 1, the result is that number
itself. However, when one divides any number by a "small" number (a
number less than 1), the result is a large number (a number greater than
1). And, contrariwise, when one divides any number by a "large" number,
the result is a "small" number.

Note that the reciprocal of zero is so large that it cannot be defined
as a number. It is said to be undefined or *infinity*, and denoted with
the symbol $\infty$.

In brief, any number (positive or negative) divided by zero is
undefined. That is: $x/0 = \infty$, where $x$ may be any number,
positive, negative, or zero.

Finally, note that when the reciprocal of a number $a$ multiplies a
number $x$, the result is the same as $x$ divided by $a$. That is

$$\left( \frac{1}{a} \right) \times x = \frac{x}{a}.$$

For example: 
$$\left( \frac{1}{2}\right) \times 10 = \frac{10}{2} = 5.$$

As an exercise, fill out all the blanks in the following spreadsheet.
Try to do the computations manually and memorize as much as you can.
Show me your completed worksheet at the beginning of the next class.

   $x$   $\frac{1}{x}$   $\frac{1}{x}$ (as a fraction)   $\frac{1}{x}$ (in decimals \[1\])    $\frac{1}{x}$ (in percentage)
  ----- --------------- ------------------------------- ----------------------------------- -------------------------------
    0      $\infty$                   \-                                \-                                               \-
   .1         10                $\frac{10}{1}$                         10.0                                         1,000 %
   .2          5                                                                                                       500%
               4                 $\frac{4}{1}$                                                                           \%
   .5          2                                                                                                         \%
    1                                                                   1.0                                            100%
    2                            $\frac{1}{2}$                          .5                                              50%
    3                            $\frac{1}{3}$                      $.\bar{3}$                                $33.\bar{3}$%
    4                                                                                                                   25%
    5                                                                                                                    \%
    6                                                                                                                    \%
    7                                                                                                                    \%
    8                                                                                                                    \%
    9                                                                                                                    \%
   10                           $\frac{1}{10}$                                                                          10%
   11                                                                                                                    \%
   12                                                                                                                    \%
   13                                                                                                                    \%

R code
======

```{r}
x <- c(0, .1, .2, .25, .5, 1:13) 
1/x 
```


Simple average
==============

Often there are data on a variable and one needs to find its *typical*,
*representative*, or **average** value. This is an example in which one
uses the simplest average concept, that of the *arithmetic mean* or
simple average.

A group of four students found in their pockets the following amounts of
cash (in dollars): $\{25, 15, 10, 30\}$. What is the amount of cash in
the pocket of a typical student in the group? Note that no particular
individual needs to have exactly that amount. One may have learned in
middle school how to calculate the simple or arithmetic average of the
data given: 
$$\frac{25 + 15 + 10 + 30}{4} = \frac{80}{4} = 20.$$

This result may be interpreted as: "The average amount of cash in the
pocket of an individual in this group is \$20."

More generally, let $x$ be any variable of interest for which one has
data $\{x_1, x_2, \ldots, x_n\}$, where $n$ is the number of values of
$x$. In statistics, $n$ is called the "sample size" or the "number of
observations" of the variable $x$. Now, let $\bar{x}$ be the simple or
arithmetic average of the data (a.k.a. "arithmetic mean"). Then:

$$\bar{x} = \frac{x_1 + x_2 + \ldots + x_n}{n}$$

If one defines $\sum_{i=1}^n x_i \equiv x_1 + x_2 + \ldots + x_n$, the
formula can be simplified to: 
$$\bar{x} = \frac{\sum_{i=1}^n x_i}{n}.$$

In general, the **summation operator** is an abbreviation used to
express the addition of numbers. If $\{x_i: i=1, 2, \ldots, n\}$ is a
sequence of $n$ numbers, then the summation of the $n$ numbers is
defined as: 
$$\sum_{i=1}^n x_i \equiv x_1 + x_2 +\cdots + x_n.$$

Thus, another way of presenting the same formula is this:
$$\bar{x} = \Big(\frac{1}{n}\Big) \sum_{i=1}^n x_i.$$

This reads as: "The simple average or mean of $x$ is the sum of the data
values of $x$, from the first to the last value, divided by the sample
size (or multiplied by the reciprocal of the sample size)."

R code
======

```{r}
x <- c(25, 15, 10, 30) 
mean(x)
```

Weighted average
================

Often, one needs to take averages of averages.

To illustrate this, call the group in the previous section's example,
group $x$. One knows that their average pocket cash is $\bar{x}_1 = 20$
dollars and that the number of people in the group is $n = 4$
individuals. Suppose now that there is a second group of people, group
$y$, with the following pocket cash (also in dollars): $\{5, 17, 8\}$.
Clearly, $\bar{y} = (5 + 17 + 8)/3 = 30/3 = 10$ dollars. Also, the
sample size of the second group is $m = 3$.

Now suppose that only the averages of each group are given:
$\bar{x} = 20$ and $\bar{y} = 10$ as well as the number of people in
each group: $n = 4$ and $m=3$, and one needs to find the typical value
for both groups taken together. Let us call $z$ the group formed by
combining the two groups.

It should be clear that one cannot just take the average of the simple
averages: $(20 + 10)/2 = 15$. Why not?

Because that would give the same "weight" to each group average in
determining the average of averages. But, the first group has *four*
people while the second group only has *three*. As a result, each
individual in the second group would be given more importance in
determining the mean of the supergroup $z$.

The correct mean requires that each individual be given the same
importance or "weight" regardless of the group in which she belongs. In
other words, since the total number of people in the two groups combined
is $n + m = 4 + 3 = 7$, then group $1$ should have a "weight" of
$n/(n+m) = 4/7$ while group $2$ should have a "weight" of
$m/(n+m) = 3/7$ in determining the average of averages.

That is, 
$$\bar{z} = (4/7) \times 20 + (3/7) \times 10 = 15.7.$$

This is called the **weighted mean** or weighted average of the two
group averages.

Note that the sum of the weights equals $1$, because $n + m$ is the
sample size of the supergroup $z$.

One can verify that this is the correct or "fair" way of computing the
average of averages by combining the two groups into a single pool of
people and taking the average of their pocket cash:
$\{25, 15, 10, 30, 5, 17, 8\}$. The overall mean of all the individuals
in the two groups now merged in a single supergroup is:
$$\frac{25 + 15 + 10 + 30 + 5 + 17 + 8}{7} = 15.7.$$

In general, the weighted average of any variable $x$ or $\bar{x}$ for a
$m$ (finite) groups is given by: 
$$\bar{x} = \omega_1 \ \bar{x}_1 + \omega_2 \ \bar{x}_2 + \dots + \omega_m \ \bar{x}_m,$$

or, more compactly: 
$$\bar{x} = \sum_{k=1}^m \omega_k \ \bar{x}_k,$$

where $\omega_k = n_k/n$ is the *weight* of group $k$ given by the
sample size for group $k$ as a fraction of the entire merged sample,
such that 
$$\sum_{k=1}^m \omega_k = 1.$$

In three towns, the average price of oranges (in \$/per bag) is,
respectively: $\bar{p}_1=4, \bar{p}_2=2, \bar{p}_3=6$. The population in
each town (in thousands) is, respectively: $n_1=12, n_2=14, n_3=18$.
Therefore, the weights for each town are: 
$$\sum_{i=1}^3 n_i = 12 + 14 + 18 = 44, $$
$$\omega_1 = \frac{n_1}{\sum_{i=1}^3 n_i} = \frac{12}{44} = 0.27, $$
$$\omega_2 = \frac{n_2}{\sum_{i=1}^3 n_i} = \frac{14}{44} = 0.32, $$
$$\omega_1 = \frac{n_1}{\sum_{i=1}^3 n_i} = \frac{18}{44} = 0.41.$$

Verify that $\sum_{i=1}^3 \omega_i = 0.27 + 0.32 + 0.41 = 1.00$. The
mean for the three towns taken together, i.e. the weighted mean, is
given by: 
$$\bar{p} = \omega_1 \ \bar{p}_1 + \omega_2 \ \bar{p}_2 + \omega_3 \ \bar{p}_3 $$
$$\bar{p} = (0.27 \times 4) + (0.32 \times 2) + (0.41 \times 6) $$
$$\bar{p} \approx 4.18.$$

Note that it was possible to use *thousands* as the units of the sample
sizes, because the weights are the sample sizes (the population in each
town) as a fraction of the entire merged sample size (the population of
the three towns added together). In the weight formulas, the thousands
in the numerators cancel out the thousands in the denominators.

The notation used in the formulas is mixed. This should help you get
comfortable with different symbols used to denote the same mathematical
objects. Different textbooks use different notations, and sometimes the
same book has to change notation from chapter to chapter or section to
section, because there are only so many letters in the alphabet.

Make your own exercises to practice the computation of simple arithmetic
means and weighted means. Intra credit for every adequate exercise that
you may create and submit.

Another exercise that should help you expand your skill set:

Using Fred (<https://fred.stlouisfed.org/>) and Excel, determine the
mean of the the quarterly real gross domestic product per capita of the
U.S. for the period 2000 Q1-2019 Q3. Also using Excel, determine the
quarterly growth rate of real GDP per capita. Submit this for intra
credit as well.

R code
======

```{r}
(20+10)/2 
(4/7)*20+(3/7)*10 
z <- c(25, 15, 10, 30, 5, 17, 8) 
mean(z) 
p <- c(4, 2, 6) 
n <- c(12, 14, 18) 
w <- n/sum(n) 
w
sum(w*p)
```

References
==========